Summary of the things that I learned

### 2024-11-14 700-formalization.tex
Improve titles, make some sentences active instead of passive

In addition to the reading, investigating and writing mathematics that lead up to the previous chapters, this thesis project also had a formalization component. Parts of Hyland's paper were carefully written out in detail in a proof assistant and added to a library of formalized mathematics with the univalent point of view, called \textit{UniMath}.

This chapter will give an overview of what was formalized, as well as point out and evaluate a couple of design decisions that were made.

\section{Some Numbers}
The code for this project was written over the course of about 18 months, spread over \href{https://github.com/UniMath/UniMath/pulls?q=is%3Apr+author%3Aarnoudvanderleer+is%3Aclosed}{20 pull requests}: 13 with content about Hyland's paper, adding 23\,361 and removing 7\,620 lines of code, and 7 with only some missing category theoretical preliminaries, adding 3\,291 and removing 816 lines.

\section{Overview of the Formalized Material}
The material that was formalized can be subdivided into material that is introduced or described in Hyland's paper, and category theoretical preliminaries that are necessary to make the proofs in Hyland's paper work.

@@

\section{Equality, Isomorphisms and Equivalences}
One of the lessons during this project was that for formalizing, it is always important to choose te right equality for the job. For example, for two categories $ \C $ and $ \D $, one can aim to prove either that $ \C = \D $ or that we have a functor $ F : \C \to \D $ that has a right and left inverse, or that we have a functor $ F : \C \to \D $ that is an adjoint equivalence. Even though for univalent categories, these three notions all coincide, in practice it is really hard to show equality of categories directly: To show $ \C = \D $, we would need to show that we have an equality $ H : \C_0 = \C_0^\prime $, and then that for all $ c, c^\prime : \C $
\[ (\C(\mathrm{transport}_H(c), \mathrm{transport}_H(c^\prime))) = \D(c, c^\prime). \]
It is usually much easier to show that there exists a functor $ F $ with an inverse, because then we do not have to show these equalities of types, but can suffice with maps between the types and equalities of the objects and morphisms that are mapped. However, for the morphisms we still have transports.

@@

\subsection{The categories in question}\label{subsec:formalized-categories}

@@

\subsection{Univalence}
All of the categories in the diagram above are univalent. The proofs of this proceed by reducing to known or easy cases. For example, we know that a full subcategory of a univalent category is again univalent because it inherits isomorphisms and equalities. Also, we know that a product of univalent categories is univalent again, since its isomorphisms and equalities are equivalent to pairs of isomorphisms and equalities of its factors. We already knew that $ \SET $ is univalent, and in this project, the category of sets, indexed over a type, was constructed and shown to be univalent.

@@

During this project, a small lemma was added to the library, showing that for a full subcategory $ \int_C D \subseteq C $, where $ D_c $ is a mere proposition for all $ c : C $, if for all $ (X, \bar X), (Y, \bar Y) : \int D $, we have $ D_{X \times Y} $ (the $ C $-product of two objects in $ \int D $ is again in $ \int D $), then $ D $ creates binary products. Also, a lemma was added that shows that if a displayed category $ D $ over $ C $, and a displayed category $ E $ over $ D $, both create binary products, then $ \sigma_D E $ also creates binary products.

Note that for many categories, the displayed morphisms are mere propositions. In particular, this holds for the `\dots data' categories of this project. For example, an algebra morphism is a function that `respects the operation', and respecting the operation is a mere proposition. In these categories, it is not necessary to give a full proof that the object $ (X \times Y, \overline{X \times Y}) $ and the morphisms $ (\pi_i, \overline{\pi_i}) $ form a binary product, because it suffices to show that for all $ C, \bar C, f, g $ and all $ \bar f : \bar C \to_f \bar A $ and $ \bar g : \bar C \to_g \bar B $, we can lift the product morphism to
\[ \overline{\langle f, g \rangle} : \bar C \to_{\langle f, g \rangle} \overline{A \times B} \]
For example, if the function $ f $ commutes with the algebra actions on $ \bar C $ and $ \bar A $, and $ g $ commutes with the actions on $ \bar C $ and $ \bar B $, then we need to show that the product morphism $ \langle f, g \rangle $ commutes with the actions on $ \bar C $ and $ \overline{A \times B} $. A lemma showing that it suffices to show that we can lift the product morphism was added to the library, and used to show that the `\dots data' displayed categories create binary products.

@@

\subsection{Chicken or egg?}
The formalization started with definitions for the objects and morphisms in question: algebraic theories and algebras. Using these, the categories were defined directly. However, since showing univalence is much easier when working with displayed categories, the definitions of the categories were decoupled from the definitions of the objects and morphisms, and instead were constructed as displayed categories. This meant that of every definition, a part was duplicated: once for the object (and morphism) types, and once for the objects and morphisms in the displayed categories. This meant that it was in theory possible to get a mismatch between, for example, the definition of algebraic theories and their category.

@@

In the end, the solution was to first define the category, and then the object and morphism types as the objects and morphisms of the category. The coercions can then be defined on the standalone object and morphism types, which works very well.

\subsection{A product of categories}\label{subsec:displayed-product}
Given two categories $ C $ and $ C^\prime $, their product $ C \times C^\prime $ can be viewed as a displayed category over $ C $ (or $ C^\prime $), where the objects over any object are the objects of $ C^\prime $, and the displayed morphisms $ g: c \to_f c^\prime $ are the morphisms $ g: C^\prime(c, c^\prime) $. There are two ways to formalize this.

@@

The second approach is by constructing the displayed category $ D $ directly, setting $ D_c = C^\prime $ and $ c \to_f c^\prime = C^\prime(c, c^\prime) $. This is slightly more work, because we do it in an elementary way instead of using category theoretical machinery. In return, this approach gives cleaner definitions in practice for the identity and composition morphisms of the total category. Therefore, halfway during the project, a switch was made from the first to the second approach.

@@

Therefore, in this project, the type of every one of these components is given a name, ending in \texttt{\_ax}, for `axiom'. For example, there are \texttt{var\_ax} and \texttt{subst\_ax} for algebraic theories, \texttt{mor\_var\_ax} and \texttt{mor\_subst\_ax} for their morphisms, \texttt{action\_ax} for algebras and \texttt{app\_ax} and \texttt{abs\_ax} for $ \lambda $-theories. The definition of the displayed categories, the constructors, the accessors and the definitions of new objects can then refer back to this.

@@

Another drawback of this approach is that for algebras and presheaves, there are in fact two different axioms. This is because their categories are displayed over the categories of algebraic theories, and the category of $ T $-algebras and $ T $-presheaves are fibers of the full displayed category. Therefore, the morphisms of $ T $-algebras are given by displayed morphisms over $ \id T $, so the axiom of an algebra morphism $ g $ is
\[ \mathtt{mor\_action\_ax}: \prod_n \prod_{t : T_n} \prod_{a : A^n} g (t \bullet a) = \id T(t) \bullet (g (a_i))_i, \]
and in practice, this makes it harder to work with them. There was an instance where one conversion from $ \id T(t) $ to $ t $ added multiple seconds to the compilation time. Therefore, for the algebras and presheaves, some of these axioms have to be stated twice: once to define the displayed category, and once for most of the other occurrences.

@@

Very early on in the project, a choice was made to use the second approach. Since many definitions require extending a mapping to a tuple, and since we do not very often need to extend tuples or do induction on a tuple, this proved to be the right choice.

@@

In this case, somewhat of a compromise was made. The endomorphism theory was indeed defined using a repeated binary product, and the construction of finite powers from the terminal object and binary products\footnote{Unfortunately, defining the product of a tuple of objects was much harder than defining the finite power of one object. When generalizing the approach to tuples of objects, some terms suddenly do not have the `correct' type, which has something to do with some equalities like $ ((x_1, \dots, x_n) + (y))_{n + 1} = y $ that do not hold definitionally. Even though this specific problem can be solved using transports, the rest of the proof becomes then much harder because the transports get in the way}. On the other hand, the Yoneda lemma for $ L $-presheaves does indeed use the `usual' notion of products $ L^n $ where the sets are tuple types. Luckily, the library contains a proof that two products are isomorphic, so where the definitions clash, we use the isomorphisms to translate between them. Even though this adds friction, in the end it is pretty straightforward.

Also, this project added a lemma to the library stating that in any category $ C $, $ \prod_{i : \emptyset} X_i $ (for $ X : \emptyset \to T $ a family of objects) is given by the terminal object. However, the repeated binary product construction needed $ \prod_{i : \{ i \mid 1 \leq i \leq 0 \}} X_i $, and $ \emptyset $ is not definitionally equal to $ \{ i \mid 1 \leq i \leq 0 \} $. Initially, this was solved by transporting along the equality $ \emptyset = \{ i \mid 1 \leq i \leq 0 \} $. However, much further down the line, some proof involved an element of this product over $ \{ i \mid 1 \leq i \leq 0 \} $, and because of the transport, rocq was not able to see that this product was just $ \{ \star \} $. Therefore, the statement of the lemma about the empty product was changed to ``For a type $ I $ and a family $ X : I \to C $, if we have a function $ f : I \to \emptyset $, the terminal object gives the product $ \prod_{i : I} X_i $''. Note that having the function $ f $ is equivalent to $ I $ being equivalent (and therefore equal) to the empty type. This change in the statement did provide the right amount of generality to make computation down the line much smoother.

\section{The \texorpdfstring{$ n + p $}{n + p}-presheaf}
Now, the definition of the endomorphism $ \lambda $-theory $ E_{\Pshf L}(L) $ requires that the theory presheaf $ L $ is exponentiable. As shown before, $ P^L $ is given by $ A(P, 1) $, the `plus 1' presheaf, where the last variable is ignored in the $ L $-action. In fact, one can show that $ P^{L^n} $ is given by $ A(P, n) $ and this was the level of generality at which the statement was formalized. To obtain the action
\[ \bullet : A(P, p)_m \times L_n^m \to A(P, p)_n, \]
the equivalence
\[ \{ 1, \dots, n + p \} \simeq \{ 1, \dots, n \} \sqcup \{ 1, \dots, p \} \]
from the library was used, which sends the first $ n $ elements to the first set, and the last $ p $ elements to the second, so this could indeed be used to construct elements of tuples $ A^{n + p} $, separating them into first a tuple of $ n $ and then a tuple of $ p $. After this, it was time to construct the presheaf morphisms for $ \rho $ and $ \lambda $ between $ L $ and $ A(L, 1) $, but this turned out to be much harder than expected, because $ A(L, 1)_n = L_{n + 1} $, whereas $ \rho $ and $ \lambda $ are functions between $ L_n $ and $ L_{S n} $, where $ S n = 1 + n $. Unfortunately, $ 1 + n $ is not definitionally equal to $ n + 1 $, and even though we might transport along a proof $ h: 1 + n = n + 1 $, this will give problems down the road.

@@

It is possible that constructing a new equivalence
\[ \{ 1, \dots, p + n \} \simeq \{1, \dots, n\} \sqcup \{1, \dots, p\}, \]
makes it possible to generalize the result to general $ p $ again. However, this equivalence would be somewhat weird, because the order of $ n $ and $ p $ are reversed. Also, proving things about these standard finite sets is quite hard, and outside the scope of this project, so this was not attempted in this project, but potentially left for future work.

\section{Quotients}\label{sec:quotients}
This project used quotients twice, once explicitly and once implicitly. The first occurrence of quotients was the construction of the `extension of scalars' functor $ f_*: \RAct N \to \RAct M $ from Lemma \ref{lem:scalar-extension}. Here, the formalization worked directly with quotients over a relation. However, the relation given by $ R : (x n, m) \sim (x, f(n) \cdot m) $ for all $ a $, is reflexive and transitive, but not necessarily symmetric. Therefore, first the `equivalence closure' $ \bar R $ of $ R $ is taken: the smallest transitive and symmetric relation that contains $ R $. This is then used for the quotient. To construct functions from this, like, the right action of the monoid $ M^\prime $, we use the universal property of the quotient: a function from $ X \times M^\prime $ that sends `related' elements $ (x_1, m_1) $ and $ (x_2, m_2) $ to the same element. However, the quotient is taken over $ \bar R $ instead of $ R $, so it is no longer possible to assume that $ x_2 = x_1 n $ and $ m_1 = f(n) \cdot m_2 $ for some $ n $. In the case of the right $ M^\prime $-action, it sufficed to add the following lemma to the library
\begin{lemma}
  For a relation $ S $ on a set $ A $ and a relation $ T $ on a set $ B $, if for $ f: A \to B $, we have that $ S(a_1, a_2) $ implies $ T(f(a_1), f(a_2)) $ for all $ a_1, a_2 : A $, then $ \bar S(a_1, a_2) $ implies $ \bar T(f(a_1), f(a_2)) $ for all $ a_1, a_2 $.
\end{lemma}
However, overall, working with these quotients in an elementary way was quite a hassle.

The second occurrence of quotients was in the proof of Corollary \ref{cor:karoubi-presheaf}, or the lemma that it is based on. Initially, the statement of the lemma was only about $ \SET $, stating that $ \iota_C \bullet - : [\bar C, \SET] \to [C, \SET] $ was an adjoint equivalence. The proof used coequalizers of functors, like in Lemma \ref{lem:karoubi-equivalence}. Note that coequalizers of set-valued functions are given pointwise by coequalizers in $ \SET $, which are formalized using quotients. Because of this, when simplifying, rocq usually reduced the goal to elementary statements about quotients, which somewhat obfuscated the relatively simple nature of the construction. Luckily, the construction could be made agnostic to the specific implementation of coequalizers, by introducing a variable
\[ \mathtt{H : Coequalizers HSET} \]
and using this instead of our known implementation. Since the proofs did not use any elements of the sets in question, it was then trivial to generalize to any category with coequalizers. It turned out that the construction was a special version of left Kan extension, which uses general colimits, and using the existing definition of Kan extension, the proof was made considerably shorter.

@@

The generalization from $ \SET $ to an arbitrary category with coequalizers, which we saw in the case of Karoubi envelopes, can also be done here, but it would take considerably more effort. The problem is that the definitions of monoids and sets with a monoid action really do use the elements of their underlying sets. For example, a monoid $ M $ has a unit $ u $ and multiplication $ \cdot $, such that $ u \cdot m = m = m \cdot u $ and $ m \cdot (n \cdot l) = (m \cdot n) \cdot l $ for all $ m, n, l : M $. The generalization for an object $ M $ in some category $ C $, turns the unit into a function $ u : I \to M $ from the terminal object and the multiplication into a function $ \mu : M \times M $ from the binary product
\footnote{Note that one does not need the full power of binary products or terminal objects to define internal monoids. A category with the product-like structure $ \otimes $ and terminal-like object $ I $ that one needs to define internal monoids, is called a \index{category!monoidal}\textit{monoidal category}.},
such that the following diagrams commute:
\begin{center}
  \begin{tikzcd}
    M \times (M \times M) \arrow[rr, "{\sim}"] \arrow[d, "{\id M \times \mu}"'] & & (M \times M) \times M \arrow[d, "{\mu \times \id M}"]\\
    M \times M \arrow[rd, "{\mu}"'] & & M \times M \arrow[ld, "{\mu}"]\\
    & M
  \end{tikzcd}
  \begin{tikzcd}
    I \times M \arrow[r, "u \times \id M"] \arrow[rd, "\sim"'] & M \times M \arrow[d, "\mu"] & M \times I \arrow[l, "\id M \times u"'] \arrow[ld, "\sim"]\\
    & M
  \end{tikzcd}
\end{center}
Such a generalization of a monoid to a general category is called an \index{monoid!internal}\textit{internal monoid}. In a similar way, sets with a monoid action can be generalized. However, this all would be a lot of work, and elementary reasoning about internal monoids of $ \SET $ is still much more cumbersome than for the ordinary set-based monoids. Also, for the fundamental theorem we only need the ordinary set-based monoids. Therefore, this generalization was not pursued any further.

\section{The Formalization of the \texorpdfstring{$ \lambda $}{lambda}-calculus} \label{sec:axiomatic-lambda-calculus}
One of the important classes of objects in this thesis is the class of $ \Lambda $-algebras: algebras for the initial $ \lambda $-theory $ \Lambda $, corresponding to the `pure' $ \lambda $-calculus. Therefore, much of this project required the pure $ \lambda $-calculus to be formalized. Now, in this thesis, it is defined as a quotient of an inductive type. However, in the UniMath library, the use of inductive types (other than a handful of basic ones like \texttt{empty}, \texttt{bool}, \texttt{nat} and \texttt{coprod}) is prohibited\footnote{This is because allowing inductive types requires a much larger trusted codebase of the proof assistant, and because most of mathematics can be formalized with $ \sum $-types or using initial algebras of functors, instead of inductive and record types.}. Also, as we saw in section \ref{sec:quotients}, working with quotients can be somewhat complicated.

@@

\subsection{Propagation of substitution}
Once the pure $ \lambda $-calculus is defined, it is not very hard to give it a $ \lambda $-theory structure, using the induction principle a couple of times to show that some identities are satisfied.

@@

Therefore, a tactic was added to the project. A first version of this tactic was a variation of
\begin{lstlisting}
Ltac reduce_lambda := (
  rewrite subst_var +
  rewrite subst_l_var +
  ...
  rewrite beta_equality
).
\end{lstlisting}
attempting to rewrite (once) with at least one, but possible multiple of the equalities. The statement \texttt{repeat reduce\_lambda} sometimes took a couple of seconds, but this saved a lot of manual work. Still, there was much room for improvement.

>>>>

In addition to the reading, investigating and writing mathematics that lead up to the previous chapters, this thesis project also had a formalization component. I carefully worked out parts of Hyland's paper in detail using a proof assistant and added the resulting code to a library of formalized mathematics with the univalent point of view, called \textit{UniMath}.

This chapter will give an overview of the mathematics that I formalized, as well as point out and evaluate a couple of design decisions that I made.

\section{Some Numbers}
I wrote the code for this project over the course of about 18 months, spread over \href{https://github.com/UniMath/UniMath/pulls?q=is%3Apr+author%3Aarnoudvanderleer+is%3Aclosed}{20 pull requests}: 13 with content about Hyland's paper, adding 23\,361 and removing 7\,620 lines of code, and 7 with only some missing category theoretical preliminaries, adding 3\,291 and removing 816 lines.

\section{Overview of the Formalized Material}
The material that I formalized can be subdivided into material that is introduced or described in Hyland's paper, and category theoretical preliminaries that are necessary to make the proofs in Hyland's paper work.

@@

\section{Equality, Isomorphisms and Equivalences}
One of the lessons that I learned during this project was that for formalizing, it is always important to choose te right equality for the job. For example, for two categories $ \C $ and $ \D $, one can aim to prove either that $ \C = \D $ or that we have a functor $ F : \C \to \D $ that has a right and left inverse, or that we have a functor $ F : \C \to \D $ that is an adjoint equivalence. Even though for univalent categories, these three notions all coincide, in practice it is really hard to show equality of categories directly: To show $ \C = \D $, we would need to show that we have an equality $ H : \C_0 = \C_0^\prime $, and then that for all $ c, c^\prime : \C $
\[ (\C(\mathrm{transport}_H(c), \mathrm{transport}_H(c^\prime))) = \D(c, c^\prime). \]
It is usually much easier to show that there exists a functor $ F $ with an inverse, because then we do not have to show these equalities of types, but can suffice with maps between the types and equalities of the objects and morphisms that are mapped. However, for the morphisms we still have transports.

@@

\subsection{The Categories in Question}\label{subsec:formalized-categories}

@@

\subsection{Univalence}
All of the categories in the diagram above are univalent. The proofs of this proceed by reducing to known or easy cases. For example, we know that a full subcategory of a univalent category is again univalent because it inherits isomorphisms and equalities. Also, we know that a product of univalent categories is univalent again, since its isomorphisms and equalities are equivalent to pairs of isomorphisms and equalities of its factors. We already knew that $ \SET $ is univalent, and in this project, I constructed the category of families of sets, indexed over a type, and showed this category to be univalent.

@@

During this project, I added a small lemma to the library, showing that for a full subcategory $ \int_C D \subseteq C $, where $ D_c $ is a mere proposition for all $ c : C $, if for all $ (X, \bar X), (Y, \bar Y) : \int D $, we have $ D_{X \times Y} $ (the $ C $-product of two objects in $ \int D $ is again in $ \int D $), then $ D $ creates binary products. I also added a lemma showing that if a displayed category $ D $ over $ C $, and a displayed category $ E $ over $ D $, both create binary products, then $ \sigma_D E $ also creates binary products.

Note that for many categories, the displayed morphisms are mere propositions. In particular, this holds for the `\dots data' categories of this project. For example, an algebra morphism is a function that `respects the operation', and respecting the operation is a mere proposition. In these categories, it is not necessary to give a full proof that the object $ (X \times Y, \overline{X \times Y}) $ and the morphisms $ (\pi_i, \overline{\pi_i}) $ form a binary product, because it suffices to show that for all $ C, \bar C, f, g $ and all $ \bar f : \bar C \to_f \bar A $ and $ \bar g : \bar C \to_g \bar B $, we can lift the product morphism to
\[ \overline{\langle f, g \rangle} : \bar C \to_{\langle f, g \rangle} \overline{A \times B} \]
For example, if the function $ f $ commutes with the algebra actions on $ \bar C $ and $ \bar A $, and $ g $ commutes with the actions on $ \bar C $ and $ \bar B $, then we need to show that the product morphism $ \langle f, g \rangle $ commutes with the actions on $ \bar C $ and $ \overline{A \times B} $. I added a lemma showing that it suffices to construct a lift for the product morphism, and used this lemma to show that the `\dots data' displayed categories create binary products.

@@

\subsection{Chicken or Egg?}
The formalization started with definitions for the objects and morphisms in question: algebraic theories and algebras. Using these, I defined the categories directly. However, since showing univalence is much easier when working with displayed categories, I decoupled the definitions of the categories from the definitions of the objects and morphisms, and instead constructed them as displayed categories. This meant that of every definition, a part was duplicated: once for the object (and morphism) types, and once for the objects and morphisms in the displayed categories. This meant that it was in theory possible to get a mismatch between, for example, the definition of algebraic theories and their category.

@@

In the end, the best solution was to first define the category, and then the object and morphism types as the objects and morphisms of the category. The coercions can then be defined on the standalone object and morphism types, which works very well.

\subsection{A Product of Categories}\label{subsec:displayed-product}
Given two categories $ C $ and $ C^\prime $, their product $ C \times C^\prime $ can be viewed as a displayed category over $ C $ (or $ C^\prime $), where the objects over any object are the objects of $ C^\prime $, and the displayed morphisms $ g: c \to_f c^\prime $ are the morphisms $ g: C^\prime(c, c^\prime) $. There are two ways to formalize this.

@@

The second approach is by constructing the displayed category $ D $ directly, setting $ D_c = C^\prime $ and $ c \to_f c^\prime = C^\prime(c, c^\prime) $. This is slightly more work, because we do it in an elementary way instead of using category theoretical machinery. In return, this approach gives cleaner definitions in practice for the identity and composition morphisms of the total category. Therefore, halfway during the project, I switched from the first to the second approach.

@@

Therefore, in this project, I gave the type of every one of these components a name, ending in \texttt{\_ax}, for `axiom'. For example, there are \texttt{var\_ax} and \texttt{subst\_ax} for algebraic theories, \texttt{mor\_var\_ax} and \texttt{mor\_subst\_ax} for their morphisms, \texttt{action\_ax} for algebras and \texttt{app\_ax} and \texttt{abs\_ax} for $ \lambda $-theories. The definition of the displayed categories, the constructors, the accessors and the definitions of new objects can then refer back to this.

@@

Another drawback of this approach is that for algebras and presheaves, there are in fact two different axioms. This is because their categories are displayed over the categories of algebraic theories, and the category of $ T $-algebras and $ T $-presheaves are fibers of the full displayed category. Therefore, the morphisms of $ T $-algebras are given by displayed morphisms over $ \id T $, so the axiom of, for example, an algebra morphism $ g $ is
\[ \mathtt{mor\_action\_ax}: \prod_n \prod_{t : T_n} \prod_{a : A^n} g (t \bullet a) = \id T(t) \bullet (g (a_i))_i, \]
and in practice, this makes it harder to work with them. There was an instance where one conversion from $ \id T(t) $ to $ t $ added multiple seconds to the compilation time. Therefore, for the algebras and presheaves, some of these axioms have to be stated twice: once to define the displayed category, and once for most of the other occurrences.

@@

Very early on in the project, I chose to use the second approach. Since many definitions require extending a mapping to a tuple, and since we do not very often need to extend tuples or do induction on a tuple, this proved to be the right choice.

@@

In this case, I made somewhat of a compromise. I indeed defined the endomorphism theory using a repeated binary product, and the construction of finite powers from the terminal object and binary products\footnote{Unfortunately, defining the product of a tuple of objects was much harder than defining the finite power of one object. When generalizing the approach to tuples of objects, some terms suddenly do not have the `correct' type, which has something to do with some equalities like $ ((x_1, \dots, x_n) + (y))_{n + 1} = y $ that do not hold definitionally. Even though this specific problem can be solved using transports, the rest of the proof becomes then much harder because the transports get in the way}. On the other hand, for the Yoneda lemma for $ L $-presheaves, I indeed use the `usual' notion of products $ L^n $ where the sets are tuple types. Luckily, the library contains a proof that products are unique up to isomorphism, so where the definitions clash, we use the isomorphisms to translate between them. Even though this adds friction, in the end it is pretty straightforward.

Also, this project added a lemma to the library stating that in any category $ C $, $ \prod_{i : \emptyset} X_i $ (for $ X : \emptyset \to T $ a family of objects) is given by the terminal object. However, the repeated binary product construction needed $ \prod_{i : \{ i \mid 1 \leq i \leq 0 \}} X_i $, and $ \emptyset $ is not definitionally equal to $ \{ i \mid 1 \leq i \leq 0 \} $. Initially, I solved this by transporting along the equality $ \emptyset = \{ i \mid 1 \leq i \leq 0 \} $. However, much further down the line, some proof involved an element of this product over $ \{ i \mid 1 \leq i \leq 0 \} $, and because of the transport, rocq was not able to see that this product was just $ \{ \star \} $. Therefore, I changed the statement of the lemma about the empty product to ``For a type $ I $ and a family $ X : I \to C $, if we have a function $ f : I \to \emptyset $, the terminal object gives the product $ \prod_{i : I} X_i $''. Note that having the function $ f $ is equivalent to $ I $ being equivalent (and therefore equal) to the empty type. This change in the statement did provide the right amount of generality to make computation down the line much smoother.

\section{The `\texorpdfstring{$ n + p $}{n + p}'-Presheaf}
Now, the definition of the endomorphism $ \lambda $-theory $ E_{\Pshf L}(L) $ requires that the theory presheaf $ L $ is exponentiable. As shown before, $ P^L $ is given by $ A(P, 1) $, the `plus 1' presheaf, where the last variable is ignored in the $ L $-action. In fact, one can show that $ P^{L^n} $ is given by $ A(P, n) $ and this was indeed the level of generality at which I formalized the statement. To obtain the action
\[ \bullet : A(P, p)_m \times L_n^m \to A(P, p)_n, \]
I used the equivalence
\[ \{ 1, \dots, n + p \} \simeq \{ 1, \dots, n \} \sqcup \{ 1, \dots, p \} \]
from the library, which sends the first $ n $ elements to the first set, and the last $ p $ elements to the second, so this could indeed be used to construct elements of tuples $ A^{n + p} $, separating them into first a tuple of $ n $ and then a tuple of $ p $. After this, it was time to construct the presheaf morphisms for $ \rho $ and $ \lambda $ between $ L $ and $ A(L, 1) $, but this turned out to be much harder than expected, because $ A(L, 1)_n = L_{n + 1} $, whereas $ \rho $ and $ \lambda $ are functions between $ L_n $ and $ L_{S n} $, where $ S n = 1 + n $. Unfortunately, $ 1 + n $ is not definitionally equal to $ n + 1 $, and even though we might transport along a proof $ h: 1 + n = n + 1 $, such a transport will give problems down the road.

@@

It is possible that constructing a new equivalence
\[ \{ 1, \dots, p + n \} \simeq \{1, \dots, n\} \sqcup \{1, \dots, p\}, \]
makes it possible to generalize the result to general $ p $ again. However, this equivalence would be somewhat weird, because the order of $ n $ and $ p $ are reversed. Also, proving things about these standard finite sets is quite hard, and outside the scope of this project, so I did not attempted this in this project, but left it for potential future work.

\section{Quotients}\label{sec:quotients}
This project used quotients twice, once explicitly and once implicitly. The first occurrence of quotients was the construction of the `extension of scalars' functor $ f_*: \RAct N \to \RAct M $ from Lemma \ref{lem:scalar-extension}. Here, the formalization worked directly with quotients over a relation. However, the relation given by $ R : (x n, m) \sim (x, f(n) \cdot m) $ for all $ a $, is reflexive and transitive, but not necessarily symmetric. Therefore, we first take the `equivalence closure' $ \bar R $ of $ R $: the smallest transitive and symmetric relation that contains $ R $. This is then used for the quotient. To construct functions from this, like the right action of the monoid $ M^\prime $, we use the universal property of the quotient: a function from $ X \times M^\prime $ that sends `related' elements $ (x_1, m_1) $ and $ (x_2, m_2) $ to the same element gives a function from the quotient $ X \times M^\prime / \sim $. However, the quotient is taken over $ \bar R $ instead of $ R $, so it is no longer possible to assume that $ x_2 = x_1 n $ and $ m_1 = f(n) \cdot m_2 $ for some $ n $. In the case of the right $ M^\prime $-action, it sufficed to add the following lemma to the library
\begin{lemma}
  For a relation $ S $ on a set $ A $ and a relation $ T $ on a set $ B $, if for $ f: A \to B $, we have that $ S(a_1, a_2) $ implies $ T(f(a_1), f(a_2)) $ for all $ a_1, a_2 : A $, then $ \bar S(a_1, a_2) $ implies $ \bar T(f(a_1), f(a_2)) $ for all $ a_1, a_2 $.
\end{lemma}
However, overall, working with these quotients in an elementary way was quite a hassle.

The second occurrence of quotients was in the proof of Corollary \ref{cor:karoubi-presheaf}, or the lemma that it is based on. Initially, the statement of the lemma was only about $ \SET $, stating that $ \iota_C \bullet - : [\bar C, \SET] \to [C, \SET] $ was an adjoint equivalence. The proof used coequalizers of functors, like in Lemma \ref{lem:karoubi-equivalence}. Note that coequalizers of set-valued functions are given pointwise by coequalizers in $ \SET $, which are formalized using quotients. Because of this, when simplifying, rocq usually reduced the goal to elementary statements about quotients, which somewhat obfuscated the relatively simple nature of the construction. Luckily, the construction could be made agnostic to the specific implementation of coequalizers, by introducing a variable
\[ \mathtt{H : Coequalizers HSET} \]
and using this instead of our known implementation. Since the proofs did not use any elements of the sets in question, it was then trivial to generalize to any category with coequalizers. It turned out that the construction was a special version of left Kan extension, which uses general colimits, and using the existing definition of Kan extension, I was able to considerably shorten the proof.

@@

The generalization from $ \SET $ to an arbitrary category with coequalizers, which we saw in the case of Karoubi envelopes, can also be done here, but it would take considerably more effort. The problem is that the definitions of monoids and sets with a monoid action really do use the elements of their underlying sets. For example, a monoid $ M $ has a unit $ u $ and multiplication $ \cdot $, such that $ u \cdot m = m = m \cdot u $ and $ m \cdot (n \cdot l) = (m \cdot n) \cdot l $ for all $ m, n, l : M $. The generalization for an object $ M $ in some category $ C $, turns the unit into a function $ u : I \to M $ from the terminal object and the multiplication into a function $ \mu : M \times M $ from the binary product
\footnote{Note that one does not need the full power of binary products or terminal objects to define internal monoids. A category with the product-like structure $ \otimes $ and terminal-like object $ I $ that one needs to define internal monoids, is called a \index{category!monoidal}\textit{monoidal category}.},
such that the following diagrams commute:
\begin{center}
  \begin{tikzcd}
    M \times (M \times M) \arrow[rr, "{\sim}"] \arrow[d, "{\id M \times \mu}"'] & & (M \times M) \times M \arrow[d, "{\mu \times \id M}"]\\
    M \times M \arrow[rd, "{\mu}"'] & & M \times M \arrow[ld, "{\mu}"]\\
    & M
  \end{tikzcd}
  \begin{tikzcd}
    I \times M \arrow[r, "u \times \id M"] \arrow[rd, "\sim"'] & M \times M \arrow[d, "\mu"] & M \times I \arrow[l, "\id M \times u"'] \arrow[ld, "\sim"]\\
    & M
  \end{tikzcd}
\end{center}
Such a generalization of a monoid to a general category is called an \index{monoid!internal}\textit{internal monoid}. In a similar way, sets with a monoid action can be generalized. However, this all would be a lot of work, and elementary reasoning about internal monoids of $ \SET $ is still much more cumbersome than for the ordinary set-based monoids. Also, for the fundamental theorem we only need the ordinary set-based monoids. Therefore, I did not pursue this generalization any further.

\section{The Formalization of the \texorpdfstring{$ \lambda $}{lambda}-Calculus} \label{sec:axiomatic-lambda-calculus}
One of the important classes of objects in this thesis is the class of $ \Lambda $-algebras: algebras for the initial $ \lambda $-theory $ \Lambda $, corresponding to the `pure' $ \lambda $-calculus. Therefore, much of this project required the pure $ \lambda $-calculus to be formalized. Now, in this thesis, it is defined as a quotient of an inductive type. However, in the UniMath library, the use of inductive types (other than a handful of basic ones like \texttt{empty}, \texttt{bool}, \texttt{nat} and \texttt{coprod}) is prohibited\footnote{This is because allowing inductive types requires a much larger trusted codebase of the proof assistant, and because most of mathematics can be formalized with $ \sum $-types or using initial algebras of functors, instead of inductive and record types.}. Also, as we saw in section \ref{sec:quotients}, working with quotients can be somewhat complicated.

@@

\subsection{Propagation of Substitution}
Once the pure $ \lambda $-calculus is defined, it is not very hard to give it a $ \lambda $-theory structure, using the induction principle a couple of times to show that some identities are satisfied.

@@

Therefore, I added a tactic to the project. A first version of this tactic was a variation of
\begin{lstlisting}
Ltac reduce_lambda := (
  rewrite subst_var +
  rewrite subst_l_var +
  ...
  rewrite beta_equality
).
\end{lstlisting}
attempting to rewrite (once) with at least one, but possible multiple of the equalities. The statement \texttt{repeat reduce\_lambda} sometimes took a couple of seconds, but this saved a lot of manual work. Still, there was much room for improvement.
